Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2017-06-06 threshold/draft.Rnw
Added generic parallel correlation example

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case, and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than simply finding
only highly correlated vectors.

>>>>

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than only finding
highly correlated vectors.

### 2017-06-06 threshold/draft.tex
Added generic parallel correlation example

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

@@

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case, and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than simply finding
only highly correlated vectors.

@@

Arbitrary vectors

>>>>

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:14:31 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:14:31 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

@@

Some applications are interested in finding both highly correlated and
anti-correlated vectors--that is, for a given threshold $t > 0$, find column
vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
similarly to Lemma \ref{sumlemma},
\begin{equation*}
\|a_i + a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
\end{equation*}
where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
anti-correlation threshold as follows:
\begin{enumerate}
\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
$P$ that orders points in $\hat{v}_1$.
\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) :
\begin{array}{l}
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
\\
e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\end{array}
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
absolute value of the matrix $D^{\tilde{i}}$.
\end{enumerate}
The modified step 3 produces a run length $\ell$ very likely to be larger than
in the default case and step 4 above requires more arithmetic. Thus, including
anti-correlated vectors in the result is more expensive than finding
only highly correlated vectors.

@@

WRITE ME
